---
title: "Module 1: Data Ingestion & Pre-Processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Module 1: Data Ingestion & Pre-Processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(agristat)
```

## Overview

Module 1 provides the complete data ingestion and pre-processing workflow for
agricultural field trials.  The workflow consists of five steps:

1. **Import** field data from CSV, Excel, or spatial formats
2. **Declare** null and alternative hypotheses
3. **Check** statistical assumptions (normality, variance homogeneity)
4. **Transform** data if assumptions are violated
5. **Report** findings with plain-language recommendations

---

## Step 1 — Import field data

```{r import}
# Create a small example dataset
tmp <- tempfile(fileext = ".csv")
write.csv(
  data.frame(
    plot      = 1:12,
    yield     = c(3.2, 4.1, 3.8, 4.5, 2.9, 3.6,
                  4.0, 3.7, 3.5, 4.2, 3.1, 4.4),
    treatment = rep(c("A", "B", "C"), times = 4)
  ),
  tmp,
  row.names = FALSE
)

field_data <- import_field_data(tmp)
print(field_data)
attr(field_data, "agristat_metadata")
```

---

## Step 2 — Declare a hypothesis

```{r hypothesis}
h <- declare_hypothesis(
  h0    = "No significant difference in yield between treatments A, B, and C",
  h1    = "At least one treatment produces a significantly different yield",
  alpha = 0.05
)
print(h)
```

---

## Step 3 — Check statistical assumptions

```{r check}
report <- check_assumptions(
  x      = field_data$yield,
  groups = field_data$treatment,
  alpha  = h$alpha
)
print(report)
```

---

## Step 4 — Apply a transformation if needed

If the normality or variance homogeneity assumption is violated, use
`auto_transform()` to identify the best data transformation:

```{r transform}
# Simulate skewed data for demonstration
set.seed(42)
skewed <- exp(rnorm(30, mean = 1.5, sd = 0.5))

tr_result <- auto_transform(skewed, response_name = "yield")
print(tr_result)
```

The plot comparing the original and transformed distributions:

```{r transform-plot, fig.width=7, fig.height=4}
tr_result$plot
```

---

## Step 5 — Get non-parametric recommendations

If parametric assumptions cannot be met even after transformation:

```{r nonpar}
recs <- suggest_nonparametric(report, n_groups = 3L)
cat(recs, sep = "\n")
```

---

## Mendelian Segregation Test

For genetics applications, test whether observed offspring ratios match
expected Mendelian ratios:

```{r segregation}
# 3:1 monohybrid cross (dominant:recessive)
seg <- chi_square_segregation(
  observed        = c(75, 25),
  expected_ratios = c(3, 1)
)
print(seg)

# 9:3:3:1 dihybrid cross
seg2 <- chi_square_segregation(
  observed        = c(90, 30, 30, 10),
  expected_ratios = c(9, 3, 3, 1)
)
print(seg2)
```

---

## Session Info

```{r session}
sessionInfo()
```
