---
title: "Field Layout Engine: CRD, RCBD, Factorial, Split-Plot, and Latin Square Designs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Field Layout Engine: CRD, RCBD, Factorial, Split-Plot, and Latin Square Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
library(agristat)
set.seed(42L)
```

## Introduction

Module 2 of the **agristat** package provides a complete toolkit for generating
and analysing common experimental designs in agricultural research:

- **CRD** — Completely Randomised Design  
- **RCBD** — Randomised Complete Block Design  
- **Factorial** — 2-way, 3-way, and n-way factorial designs  
- **Split-Plot** — Split-plot, split-split-plot, and strip-plot designs  
- **Latin Square** — Two-way blocking with row and column control  

Every `design_*()` function returns an S3 object of class `field_design` that
stores the plot-level layout, error structure, and metadata.  The
`analyze_design()` dispatcher automatically selects the correct ANOVA model for
each design type.

---

## 1. Completely Randomised Design (CRD)

A CRD is appropriate when experimental material is homogeneous.  Treatments are
assigned to plots in a completely random order with no blocking.

```{r crd}
d_crd <- design_crd(
  treatments = c("Control", "Low-N", "High-N"),
  replicates = 4L,
  seed       = 42L
)
print(d_crd)
```

### Visualise the layout

```{r crd-plot}
plot_field_layout(d_crd, title = "CRD: Nitrogen Treatments")
```

### Simulate data and run ANOVA

```{r crd-anova}
set.seed(1L)
df_crd         <- d_crd$layout
df_crd$yield   <- c(
  rnorm(4, mean = 3.2, sd = 0.4),   # Control
  rnorm(4, mean = 4.1, sd = 0.4),   # Low-N
  rnorm(4, mean = 5.0, sd = 0.4)    # High-N
)
result_crd <- analyze_design(d_crd, response = "yield", data = df_crd)
summary(result_crd)
```

### Diagnostic plots

```{r crd-diag, fig.height = 6}
plot_anova_diagnostics(result_crd, which = 1:2)
```

---

## 2. Randomised Complete Block Design (RCBD)

RCBD is used when there is a known source of field variation (e.g., a soil
fertility gradient).  Each block contains one complete set of treatments.

```{r rcbd}
d_rcbd <- design_rcbd(
  treatments = c("V1", "V2", "V3", "V4"),
  blocks     = 4L,
  seed       = 7L
)
print(d_rcbd)
```

```{r rcbd-plot}
plot_field_layout(d_rcbd, title = "RCBD: Variety Trial")
```

```{r rcbd-anova}
set.seed(2L)
df_rcbd         <- d_rcbd$layout
block_effect    <- as.numeric(df_rcbd$block) * 0.5
trt_effect      <- c(V1 = 0, V2 = 0.8, V3 = 1.5, V4 = 2.2)[df_rcbd$treatment]
df_rcbd$yield   <- 10 + block_effect + trt_effect + rnorm(nrow(df_rcbd), 0, 0.3)

result_rcbd <- analyze_design(d_rcbd, response = "yield", data = df_rcbd)
summary(result_rcbd)
```

---

## 3. Factorial Design

A factorial design crosses all combinations of two or more factors.  All main
effects and interaction terms are estimable.

```{r factorial}
d_fact <- design_factorial(
  factor_names  = c("Nitrogen", "Variety"),
  factor_levels = list(
    Nitrogen = c("Low", "High"),
    Variety  = c("V1", "V2", "V3")
  ),
  replicates = 3L,
  seed       = 1L
)
print(d_fact)
cat("Contrast terms:", paste(d_fact$metadata$contrast_coding, collapse = ", "), "\n")
```

```{r factorial-plot}
plot_field_layout(d_fact, title = "2 x 3 Factorial Layout")
```

```{r factorial-anova}
set.seed(3L)
df_fact         <- d_fact$layout
df_fact$yield   <- with(df_fact, {
  mu <- 10
  n  <- ifelse(Nitrogen == "High", 1.2, 0)
  v  <- c(V1 = 0, V2 = 0.5, V3 = 1.0)[Variety]
  mu + n + v + n * v * 0.3 + rnorm(nrow(df_fact), 0, 0.4)
})
result_fact <- analyze_design(d_fact, response = "yield", data = df_fact)
summary(result_fact)
```

---

## 4. Split-Plot Design

Split-plot designs arise when it is impractical to randomise all factor
combinations within each block.  The whole-plot factor is applied to large
experimental units; the sub-plot factor is applied within each whole-plot.

```{r split-plot}
d_sp <- design_split_plot(
  whole_plot_factor = c("Irrigation-On", "Irrigation-Off"),
  sub_plot_factor   = c("V1", "V2", "V3"),
  blocks            = 4L,
  seed              = 99L
)
print(d_sp)
```

```{r split-plot-errors}
compute_error_terms(d_sp)
```

```{r split-plot-anova}
set.seed(4L)
df_sp         <- d_sp$layout
df_sp$yield   <- with(df_sp, {
  irr <- ifelse(whole_plot_level == "Irrigation-On", 2, 0)
  var <- c(V1 = 0, V2 = 0.8, V3 = 1.5)[sub_plot_level]
  10 + irr + var + rnorm(nrow(df_sp), 0, 0.5)
})
result_sp <- analyze_design(d_sp, response = "yield", data = df_sp)
summary(result_sp)
```

---

## 5. Latin Square Design

A Latin Square provides two-way blocking, controlling variation in two
orthogonal directions (rows and columns).  Each treatment appears exactly
once in each row and each column.

```{r latin-square}
d_ls <- design_latin_square(
  factor_levels = c("A", "B", "C", "D"),
  seed          = 2024L
)
print(d_ls)
```

```{r latin-square-plot}
plot_field_layout(d_ls, title = "4 x 4 Latin Square")
```

```{r latin-square-anova}
set.seed(5L)
df_ls         <- d_ls$layout
df_ls$yield   <- with(df_ls, {
  trt_eff <- c(A = 0, B = 0.5, C = 1.2, D = 2.0)[treatment]
  row_eff <- (as.numeric(row) - 2.5) * 0.3
  col_eff <- (as.numeric(column) - 2.5) * 0.2
  10 + trt_eff + row_eff + col_eff + rnorm(nrow(df_ls), 0, 0.3)
})
result_ls <- analyze_design(d_ls, response = "yield", data = df_ls)
summary(result_ls)
```

---

## 6. Error Structure Inspection

You can inspect the theoretical error structure of any design with
`get_error_structure()`:

```{r error-structure}
# Split-plot: two error strata
es_sp <- get_error_structure(d_sp)
str(es_sp)
```

---

## 7. Strip-Plot and Split-Split-Plot

```{r strip-plot}
d_strip <- design_strip_plot(
  row_factor = c("Herbicide-A", "Herbicide-B", "Herbicide-C"),
  col_factor = c("Low-N", "Mid-N", "High-N"),
  blocks     = 3L,
  seed       = 22L
)
print(d_strip)
get_error_structure(d_strip)
```

```{r split-split-plot}
d_ssp <- design_split_split_plot(
  whole_plot_factor   = c("Tillage-Deep", "Tillage-Shallow"),
  sub_plot_factor     = c("Irrigation-On", "Irrigation-Off"),
  sub_sub_plot_factor = c("V1", "V2"),
  blocks              = 3L,
  seed                = 5L
)
print(d_ssp)
names(d_ssp$error_structure)
```

---

## Summary

| Function | Design | ANOVA model |
|---|---|---|
| `design_crd()` | CRD | `aov(y ~ treatment)` |
| `design_rcbd()` | RCBD | `aov(y ~ treatment + block)` |
| `design_factorial()` | Factorial | `aov(y ~ A * B * ...)` |
| `design_split_plot()` | Split-plot | `lmer(y ~ A * B + (1\|wp))` |
| `design_split_split_plot()` | Split-split-plot | `lmer(...)` |
| `design_strip_plot()` | Strip-plot | `aov(y ~ block + A + B + A:B)` |
| `design_latin_square()` | Latin Square | `aov(y ~ row + column + treatment)` |

All designs use `analyze_design()` as a unified dispatcher, and their layouts
can be visualised with `plot_field_layout()`.
