---
title: "Module 3: Breeding & Genetics Analytics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Module 3: Breeding & Genetics Analytics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Overview

Module 3 of **agristat** provides tools for breeding programme design and
quantitative genetics analysis:

- **Experimental designs**: alpha-lattice and augmented designs for large
  multi-genotype trials
- **GxE analysis**: multi-environment trial (MET) decomposition
- **AMMI**: additive main effects and multiplicative interaction modelling
- **Heritability**: broad- and narrow-sense estimates with bootstrap CIs
- **QTL mapping**: data preparation and LOD visualisation
- **GWAS**: genome-wide association with Manhattan and Q-Q plots

## Experimental Designs

### Alpha-Lattice Design (large trials)

An alpha-lattice design is a resolvable balanced incomplete block design
suitable for evaluating 100+ genotypes.  Each replicate contains every
genotype exactly once, distributed across equal-sized incomplete blocks.

```{r alpha-lattice}
library(agristat)

# 100 genotypes, blocks of 10, 3 replicates
design_al <- design_alpha_lattice(
  num_genotypes = 100L,
  block_size    = 10L,
  replicates    = 3L,
  seed          = 2024L
)

head(design_al)
nrow(design_al) # 300 plots total
```

### Augmented Design (unreplicated test lines + replicated checks)

```{r augmented}
test_lines  <- paste0("Line", sprintf("%03d", 1:200))
check_vars  <- c("Std_Check_1", "Std_Check_2", "Std_Check_3")

design_aug <- design_augmented(
  test_genotypes    = test_lines,
  check_genotypes   = check_vars,
  replicates_checks = 5L,    # 5 blocks
  seed              = 101L
)

table(design_aug$type)
```

## GxE Interaction Analysis

```{r gxe}
# Simulate multi-environment trial
set.seed(1L)
met_data <- expand.grid(
  genotype    = paste0("G", 1:10),
  environment = paste0("E", 1:6),
  stringsAsFactors = FALSE
)
# Add simulated yield with GxE
met_data$yield <- rnorm(nrow(met_data), mean = 5, sd = 1) +
  rep(c(0.5, -0.5, 0.2, -0.2, 0.3, -0.3), each = 10)

gxe_res <- analyze_gxe(met_data, response = "yield")
gxe_res
summary(gxe_res)
```

## AMMI Analysis

AMMI (Additive Main Effects and Multiplicative Interaction) decomposes the
GxE interaction matrix using SVD.

### Fitting the AMMI Model

```{r ammi-fit}
am_res <- fit_ammi(met_data, response = "yield")
print(am_res)
summary(am_res)
```

### AMMI Biplot

```{r ammi-biplot, fig.width=7, fig.height=6}
p_biplot <- plot_ammi_biplot(am_res, pc_x = 1L, pc_y = 2L)
print(p_biplot)
```

The biplot shows:
- **Blue points**: genotype scores (distance from origin indicates
  sensitivity to the environment)
- **Red arrows**: environment loadings (arrow direction indicates
  discriminating ability)

## Heritability Estimation

### Input: variance components from a REML fit

```{r heritability}
vc_df <- data.frame(
  source   = c("genotype", "additive", "residual"),
  variance = c(2.5,       1.8,       1.0)
)

h_res <- calc_heritability(vc_df, n_boot = 500L, ci_level = 0.95, seed = 42L)
print(h_res)
summary(h_res)
```

**Broad-sense heritability** \(H^2\) captures all genetic effects
(additive, dominance, epistasis).  **Narrow-sense heritability**
\(h^2\) captures only additive effects and predicts parent-offspring
resemblance.

## QTL Mapping

### Preparing Data

```{r qtl-prep}
set.seed(200L)
n_ind  <- 100L
n_mrk  <- 50L

pheno_df <- data.frame(
  id    = paste0("F2_", seq_len(n_ind)),
  yield = rnorm(n_ind, mean = 8, sd = 1.5)
)

marker_mat <- matrix(
  sample(0:2, n_ind * n_mrk, replace = TRUE),
  nrow     = n_ind,
  dimnames = list(paste0("F2_", seq_len(n_ind)),
                  paste0("M", seq_len(n_mrk)))
)

qtl_input <- prepare_qtl_data(pheno_df, marker_mat)
str(qtl_input)
```

### Visualising LOD Scores

```{r qtl-plot, fig.width=8, fig.height=4}
# Simulate a LOD scan result
lod_scan <- data.frame(
  chr = rep(1:5, each = 20L),
  pos = rep(seq(0, 95, 5), 5L),
  lod = abs(rnorm(100L, mean = 1, sd = 1.5))
)

p_qtl <- plot_qtl_map(lod_scan, threshold = 3.0, trait_name = "Grain Yield")
print(p_qtl)
```

## GWAS Pipeline

### Running the Association Analysis

```{r gwas-run}
set.seed(300L)
n_ind  <- 200L
n_snps <- 500L

pheno_gwas <- rnorm(n_ind, mean = 10, sd = 2)

snp_matrix <- matrix(
  sample(0:2, n_ind * n_snps, replace = TRUE,
         prob = c(0.25, 0.50, 0.25)),
  nrow     = n_ind,
  dimnames = list(NULL, paste0("SNP", seq_len(n_snps)))
)

snp_positions <- data.frame(
  marker     = paste0("SNP", seq_len(n_snps)),
  chromosome = rep(1:10, each = 50L),
  position   = rep(seq(1e5, 5e6, length.out = 50L), 10L)
)

gwas_out <- run_basic_gwas(
  phenotype       = pheno_gwas,
  marker_genotypes = snp_matrix,
  marker_positions = snp_positions,
  trait_type       = "quantitative",
  min_maf          = 0.05
)

print(gwas_out)
summary(gwas_out)
```

### Manhattan Plot

```{r manhattan, fig.width=10, fig.height=5}
p_man <- plot_manhattan(
  gwas_out,
  title   = "GWAS: Grain Yield",
  top_n   = 5L
)
print(p_man)
```

### Q-Q Plot (Genomic Inflation)

```{r qq-plot, fig.width=6, fig.height=6}
p_qq <- plot_qq_gwas(gwas_out, title = "Q-Q Plot: Grain Yield GWAS")
print(p_qq)
```

A Q-Q plot close to the diagonal indicates the null hypothesis holds for
most markers (no inflation).  The genomic inflation factor \(\lambda\)
should be close to 1.0 under a well-controlled study.

## Key Concepts

| Concept | Interpretation |
|---|---|
| \(H^2 \geq 0.8\) | High heritability: selection is highly effective |
| AMMI IPCA1 > 70% | Single multiplicative term explains most interaction |
| LOD > 3.0 | Genome-wide significant QTL |
| GWAS \(-\log_{10}(p) > 7.3\) | Bonferroni-significant with 50 million SNPs |
| \(\lambda > 1.05\) | Possible population structure inflation |

## References

- Gauch, H. G. (1988). Model selection and validation for yield trials.
  *Biometrics*, 44, 705--715.
- Falconer, D. S. & Mackay, T. F. C. (1996). *Introduction to Quantitative
  Genetics* (4th ed.). Longman.
- Visscher, P. M. et al. (2017). 10 years of GWAS discovery.
  *American Journal of Human Genetics*, 101, 5--22.
- Federer, W. T. (1956). Augmented designs. *Hawaiian Planters's Record*,
  55, 191--208.
