---
title: "04 — Mean Separation & Post-Hoc Tests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{04 — Mean Separation & Post-Hoc Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r load, message=FALSE}
library(agristat)
```

## Introduction

After a significant ANOVA F-test, the practitioner needs to determine *which*
treatment means differ. This is the **mean separation** or **post-hoc testing**
step. The `agristat` package provides four methods:

| Method | Function | Error-rate control |
|--------|----------|--------------------|
| Duncan's MRT | `dmrt()` | Experimentwise |
| Tukey's HSD | `tukey_hsd()` | Family-wise (FWER) |
| Dunnett's test | `dunnett_test()` | Family-wise vs control |
| Orthogonal contrasts | `test_contrasts()` | Per-contrast |

The unified dispatcher `mean_separation()` routes to the appropriate function
automatically.

---

## Example Dataset

We use a simple one-way ANOVA with four rice varieties:

```{r data}
set.seed(2024)
dat <- data.frame(
  yield = c(
    rnorm(5, mean = 4.2, sd = 0.3),   # Variety A
    rnorm(5, mean = 3.8, sd = 0.3),   # Variety B
    rnorm(5, mean = 5.1, sd = 0.3),   # Variety C
    rnorm(5, mean = 4.5, sd = 0.3)    # Variety D
  ),
  variety = factor(rep(c("A", "B", "C", "D"), each = 5))
)

fit <- aov(yield ~ variety, data = dat)
summary(fit)
```

The ANOVA is significant. We can now proceed to mean separation.

---

## 1. Duncan's Multiple Range Test (DMRT)

DMRT is widely used in agricultural journals. It maximises power (most
liberal) and is best for exploratory experiments.

```{r dmrt}
res_dmrt <- mean_separation(fit, method = "dmrt")
print(res_dmrt)
```

### Direct call

```{r dmrt-direct}
cell_means <- tapply(dat$yield, dat$variety, mean)
aov_sum    <- summary(fit)[[1]]
mse        <- aov_sum["Residuals", "Mean Sq"]
df_res     <- aov_sum["Residuals", "Df"]

dmrt_res <- dmrt(
  means = cell_means,
  mse   = mse,
  df    = df_res,
  alpha = 0.05,
  n     = 5
)
print(dmrt_res)
```

### Visualisation

```{r plot-dmrt, fig.cap="Treatment means with DMRT letters"}
plot_means(res_dmrt, y_label = "Grain Yield (t/ha)", mse = mse, n_rep = 5)
```

---

## 2. Tukey's Honestly Significant Difference (HSD)

Tukey's HSD controls the family-wise error rate at $\alpha$ for *all*
pairwise comparisons simultaneously. Recommended for confirmatory studies.

```{r tukey}
res_tukey <- mean_separation(fit, method = "tukey")
print(res_tukey)
```

### Interval plot

```{r plot-tukey, fig.cap="Tukey HSD pairwise confidence intervals"}
plot_tukey(res_tukey$results)
```

---

## 3. Comparing DMRT vs Tukey

```{r compare}
cat("DMRT letters:\n")
print(res_dmrt$means_with_letters[, c("treatment", "mean", "letters")])

cat("\nTukey letters:\n")
print(res_tukey$means_with_letters[, c("treatment", "mean", "letters")])
```

DMRT typically produces more significant differences (more groups) than
Tukey's HSD because it uses a lower critical value for small spans.

---

## 4. Dunnett's Test (Treatment vs Control)

When the objective is only to compare each treatment against a **control
group**, Dunnett's test provides greater power than pairwise tests.

```{r dunnett-data}
set.seed(2024)
dat2 <- data.frame(
  response = c(
    rnorm(6, mean = 10.0, sd = 1.0),   # control
    rnorm(6, mean = 12.5, sd = 1.0),   # drug A
    rnorm(6, mean = 10.3, sd = 1.0)    # drug B
  ),
  group = factor(rep(c("control", "drugA", "drugB"), each = 6))
)
fit2 <- aov(response ~ group, data = dat2)
```

```{r dunnett}
res_dunnett <- mean_separation(fit2, method = "dunnett",
                               control_group = "control")
print(res_dunnett)
```

Alternatively, call `dunnett_test()` directly:

```{r dunnett-direct}
d_res <- dunnett_test(
  response  = dat2$response,
  group     = dat2$group,
  control   = "control",
  alpha     = 0.05,
  alternative = "two.sided"
)
print(d_res)
```

---

## 5. Orthogonal Contrasts

Orthogonal contrasts partition the treatment SS into interpretable
single-degree-of-freedom components. They are best for **planned**,
hypothesis-driven comparisons.

```{r contrasts-data}
set.seed(2024)
dat3 <- data.frame(
  yield = c(
    rnorm(4, mean = 10, sd = 1),   # A: low N
    rnorm(4, mean = 12, sd = 1),   # B: medium N
    rnorm(4, mean = 14, sd = 1),   # C: high N
    rnorm(4, mean = 11, sd = 1)    # D: high N + micronutrient
  ),
  trt = factor(rep(c("A", "B", "C", "D"), each = 4))
)
fit3 <- aov(yield ~ trt, data = dat3)
```

Define a set of **orthogonal** contrasts:

```{r contrasts}
# Contrasts (rows must sum to 0; inner products must equal 0)
cmat <- rbind(
  "Low vs High N"    = c( 1,  1, -1, -1) / 2,
  "A vs B"           = c( 1, -1,  0,  0),
  "C vs D"           = c( 0,  0,  1, -1)
)

validate_contrasts(cmat)  # Should return TRUE

res_contr <- mean_separation(fit3, method = "contrasts",
                              contrast_matrix = cmat)
print(res_contr)
format_contrast_output(res_contr$results)
```

---

## 6. Choosing the Right Test

```
Is there a designated control group?
├─ YES → Use dunnett_test()  (most powerful for this case)
└─ NO
   ├─ Were comparisons planned before collecting data?
   │  ├─ YES → Use test_contrasts()  (greatest power for specific hypotheses)
   │  └─ NO (all pairwise)
   │     ├─ Confirmatory / publication with FWER control → tukey_hsd()
   │     └─ Exploratory / maximum power                  → dmrt()
```

---

## 7. Unequal Replication

Both `dmrt()` and `tukey_hsd()` handle unequal replication via the harmonic
mean approach:

```{r unequal}
set.seed(7)
dat_unequal <- data.frame(
  yield = c(rnorm(3, 10), rnorm(5, 12), rnorm(4, 11)),
  trt   = factor(c(rep("A", 3), rep("B", 5), rep("C", 4)))
)
fit_uneq <- aov(yield ~ trt, data = dat_unequal)
res_uneq  <- mean_separation(fit_uneq, method = "dmrt")
print(res_uneq$means_with_letters)
```

---

## Session info

```{r session}
sessionInfo()
```
