---
title: "Module 5: Phylogenetics & Evolutionary Mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Module 5: Phylogenetics & Evolutionary Mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(agristat)
```

## Overview

Module 5 of the **agristat** package provides a complete phylogenetic analysis
workflow:

1. **`read_fasta()`** — Import and validate DNA/protein sequences
2. **`align_sequences()`** — Multiple sequence alignment (ClustalOmega / MUSCLE)
3. **`build_tree_ml()`** — Maximum Likelihood tree construction (GTR, LG, ...)
4. **`build_tree_nj()`** — Neighbor-Joining tree construction (JC69, p-distance)
5. **`bootstrap_tree()`** — Bootstrap support values (0–100 %)
6. **`plot_phylo()`** / **`plot_tree_unrooted()`** — Tree visualisation
7. **`write_tree()`** — Export to Newick / Nexus
8. **`reroot_tree()`** — Reroot on outgroup

---

## Example DNA sequences

```{r prepare-fasta}
dna_lines <- c(
  ">AtCOI_Arabidopsis",  "ATGCATGCATGCATGCATGCATGCATGCATGCATGCATGC",
  ">OsCOI_Oryza",        "ATGCTTGCATGCATGCTTGCATGCTTGCATGCTTGCATGC",
  ">ZmCOI_Maize",        "ATGCATGCTTGCATGCATGCTTGCATGCATGCTTGCATGC",
  ">SbCOI_Sorghum",      "ATGCATGCATGCTTGCATGCATGCATGCATGCTTGCATGC",
  ">HvCOI_Barley",       "ATGCATGCATGCATGCTTGCATGCATGCATGCATGCTTGC"
)
tmp_fasta <- tempfile(fileext = ".fasta")
writeLines(dna_lines, tmp_fasta)
```

---

## Step 1: Import and validate sequences

```{r read-fasta}
seqs <- read_fasta(tmp_fasta)
print(seqs)
summary(seqs)
```

The returned object contains:

- `seqs$sequences` — named character vector of sequences
- `seqs$type` — `"nucleotide"` or `"protein"`
- `seqs$stats` — data frame with `name`, `length`, `gc_content`
- `seqs$metadata` — file, n_seqs, parse_date

```{r fasta-stats}
seqs$stats
```

---

## Step 2: Multiple sequence alignment

```{r align}
# Falls back to gap-padding when msa package is unavailable
aln <- align_sequences(seqs, method = "clustal")
print(aln)
```

```{r aln-matrix}
# Inspect first few positions of the alignment matrix
aln$alignment_matrix[, 1:10]
```

---

## Step 3a: Neighbor-Joining tree

```{r nj-tree, fig.width=6, fig.height=4}
nj_tree <- build_tree_nj(aln)
print(nj_tree)
```

Inspect the pairwise distance matrix:

```{r nj-dm}
round(nj_tree$distance_matrix, 4)
```

---

## Step 3b: Maximum Likelihood tree

```{r ml-tree}
ml_tree <- build_tree_ml(aln, model = "GTR")
print(ml_tree)
```

---

## Step 4: Bootstrap support

```{r bootstrap, fig.width=6, fig.height=4}
bs <- bootstrap_tree(
  aln,
  tree_method    = "nj",
  num_replicates = 100,
  seed           = 42
)
print(bs)
summary(bs)
```

```{r support-values}
bs$support_values
```

Interpreting bootstrap values:

| Support % | Interpretation          |
|-----------|-------------------------|
| ≥ 95      | Very strong support     |
| 70 – 94   | Moderate support        |
| 50 – 69   | Weak support            |
| < 50      | Not reliable            |

---

## Step 5: Tree visualisation

```{r plot-tree, fig.width=7, fig.height=5}
# Rooted tree with bootstrap support labels
plot_phylo(bs, main = "NJ Bootstrap Tree", support_threshold = 70)
```

```{r plot-unrooted, fig.width=7, fig.height=5}
# Unrooted (radial) display
plot_tree_unrooted(nj_tree, main = "Unrooted NJ Tree")
```

---

## Step 6: Rerooting the tree

```{r reroot}
rerooted <- reroot_tree(nj_tree, outgroup = "HvCOI_Barley")
class(rerooted)      # ape::phylo
rerooted$tip.label
```

---

## Step 7: Exporting trees

```{r write-newick}
nwk_file <- tempfile(fileext = ".nwk")
write_tree(bs, file = nwk_file, format = "newick")
cat(readLines(nwk_file), sep = "\n")
```

```{r write-nexus}
nex_file <- tempfile(fileext = ".nex")
write_tree(bs, file = nex_file, format = "nexus")
# Show first 10 lines of the NEXUS file
cat(head(readLines(nex_file, warn = FALSE), 10), sep = "\n")
```

---

## Step 8: Utility functions

### Validate sequence characters

```{r validate}
test_seqs <- c(
  Good    = "ATGCATGCATGCNRYW",
  BadChar = "ATGZ1234ATGC"
)
validate_sequence_format(test_seqs, type = "nucleotide")
```

### Extract bootstrap values

```{r extract-bs}
sv <- extract_bootstrap_values(bs)
hist(sv, main = "Bootstrap Support Distribution",
     xlab = "Support (%)", col = "steelblue", border = "white")
```

### Compute distance matrix directly

```{r dist-matrix}
dm <- compute_distance_matrix(aln$alignment_matrix,
                               seq_type = "nucleotide",
                               method   = "JC69")
round(dm, 4)
```

---

## Complete workflow example

```{r complete-workflow, eval=FALSE}
# 1. Read sequences
seqs <- read_fasta("my_sequences.fasta")

# 2. Align
aln <- align_sequences(seqs, method = "clustal")

# 3. Build tree
nj  <- build_tree_nj(aln)
ml  <- build_tree_ml(aln, model = "GTR")

# 4. Bootstrap
bs <- bootstrap_tree(aln, tree_method = "nj", num_replicates = 1000,
                     seed = 42)

# 5. Visualise
plot_phylo(bs, support_threshold = 70)

# 6. Reroot
re <- reroot_tree(bs, outgroup = "outgroup_taxon")

# 7. Export
write_tree(re, "output_tree.nwk", format = "newick")
write_tree(re, "output_tree.nex", format = "nexus")
```

---

## Package requirements

| Task | Package | Source |
|------|---------|--------|
| Multiple alignment | **msa** | Bioconductor |
| ML trees | **phangorn** | CRAN |
| NJ trees, distances | **ape** | CRAN |
| Tree visualisation | **ggtree** | Bioconductor (optional) |

Install Bioconductor packages with:
```r
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("msa", "ggtree"))
```

---

## References

- Jukes & Cantor (1969). Evolution of protein molecules.
  *Mammalian Protein Metabolism*, 3, 21–132.
- Saitou & Nei (1987). The neighbor-joining method.
  *Mol. Biol. Evol.*, 4(4), 406–425.
- Felsenstein (1981). Evolutionary trees from DNA sequences.
  *J. Mol. Evol.*, 17, 368–376.
- Felsenstein (1985). Confidence limits on phylogenies.
  *Evolution*, 39(4), 783–791.
- Schliep (2011). phangorn: phylogenetic analysis in R.
  *Bioinformatics*, 27(4), 592–593.
