---
title: "06 — Thesis Assembly & Export"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{06 — Thesis Assembly & Export}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
```

```{r load, message=FALSE}
library(agristat)
```

## Introduction

Module 6 of **agristat** automates the final step of an agricultural research
workflow: assembling thesis chapters and exporting publication-ready outputs.
The module provides:

| Function | Purpose |
|---|---|
| `generate_methods_text()` | Auto-generate a methods-section paragraph |
| `export_results_table()` | Format results as Markdown / HTML / LaTeX / CSV |
| `export_figure()` | Save ggplot2 or base-R plots in publication formats |
| `generate_appendix()` | Produce appendix with descriptive stats & assumption tests |
| `compile_report()` | Assemble all components into one Rmd/Markdown report |
| `thesis_template_methods()` | Methods-section template with fill-in placeholders |
| `thesis_template_results()` | Results-section template |
| `thesis_template_discussion()` | Discussion-section template |
| `fill_template()` | Replace `{placeholder}` tokens with actual values |

---

## 1  Shared Example Data

We use a simple rice-yield trial throughout this vignette.

```{r data}
set.seed(2024)

# Experimental design
rcbd_design <- design_rcbd(
  treatments = c("A", "B", "C", "D"),
  blocks     = 5L,
  seed       = 2024L
)

# Attach simulated yield to the randomised layout
field_data <- rcbd_design$layout
set.seed(2024)
mean_yields <- c(A = 4.2, B = 3.8, C = 4.9, D = 3.5)
field_data$yield <- mean_yields[field_data$treatment] + rnorm(nrow(field_data), 0, 0.4)

# For convenience, also keep a plain data frame with original variety labels
plain_data <- data.frame(
  yield   = field_data$yield,
  variety = field_data$treatment,
  block   = field_data$block
)

# ANOVA
aov_result <- analyze_design(
  design   = rcbd_design,
  response = "yield",
  data     = field_data
)

# DMRT post-hoc
variety_means <- tapply(plain_data$yield, plain_data$variety, mean)
dmrt_result   <- dmrt(variety_means, mse = 0.16, df = 12, n = 5)
```

---

## 2  Generating the Methods Text

`generate_methods_text()` inspects the design object and automatically
writes a methods paragraph.

```{r methods-text}
mt <- generate_methods_text(
  design_obj  = rcbd_design,
  analysis_obj = aov_result,
  posthoc_obj  = dmrt_result,
  software     = TRUE
)

cat(mt$text)
```

The object has three named sections and a references vector:

```{r methods-sections}
names(mt$sections)
mt$references
```

---

## 3  Exporting Results Tables

### 3.1  ANOVA Table

```{r anova-markdown}
cat(export_results_table(aov_result, format = "markdown"))
```

### 3.2  Means & Letter Groups

```{r means-markdown}
cat(export_results_table(dmrt_result, format = "markdown"))
```

### 3.3  HTML Table

```{r means-html, results='asis'}
cat(export_results_table(dmrt_result, format = "html"))
```

### 3.4  LaTeX Table (for inclusion in `.tex` files)

```{r means-latex}
cat(export_results_table(aov_result, format = "latex"))
```

### 3.5  CSV Export

```{r means-csv}
cat(export_results_table(dmrt_result, format = "csv"))
```

### 3.6  Writing to a File

```{r table-file, eval=FALSE}
export_results_table(
  aov_result,
  format = "markdown",
  file   = "anova_table.md"
)
```

---

## 4  Exporting Figures

### 4.1  Saving a ggplot2 Figure

```{r export-figure, eval=FALSE}
library(ggplot2)

p <- ggplot(plain_data, aes(variety, yield, fill = variety)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Grain Yield by Variety", x = "Variety",
       y = "Yield (t/ha)") +
  theme_classic()

info <- export_figure(
  plot_obj = p,
  file     = "yield_boxplot",
  format   = "png",
  width    = 7,
  height   = 5,
  dpi      = 300,
  title    = "Figure 1: Grain Yield by Rice Variety"
)

cat("Saved:", info$file_path, "\n")
cat("Size:", info$size_kb, "KB\n")
```

### 4.2  Supported Formats

| `format` | Description |
|---|---|
| `"png"` | Raster; default for on-screen and Word |
| `"pdf"` | Vector; ideal for LaTeX-based theses |
| `"svg"` | Scalable vector; good for HTML reports |
| `"eps"` | Encapsulated PostScript; required by some journals |

---

## 5  Generating an Appendix

`generate_appendix()` creates three appendix sections automatically:

```{r appendix}
app <- generate_appendix(
  data         = field_data,
  design_obj   = rcbd_design,
  analysis_obj = aov_result
)

cat("Observations:", app$n_obs, "\n")
cat("Missing values:", app$n_missing, "\n")
cat("Sections:", paste(names(app$sections), collapse = ", "), "\n")
```

### Appendix A — Data Summary

```{r appendix-data, results='asis'}
cat(app$sections$data_summary)
```

### Appendix B — Assumption Tests

```{r appendix-assump, results='asis'}
cat(app$sections$assumption_tests)
```

---

## 6  Using Thesis Templates

### 6.1  Methods Template

Templates contain `{placeholder}` tokens that you fill with real values
using `fill_template()`.

```{r template-methods}
tmpl_m <- thesis_template_methods(
  design_type    = "rcbd",
  analysis_type  = "anova",
  experiment_type = "plant"
)

# See the available placeholders
tmpl_m$placeholders
```

```{r fill-methods}
filled_methods <- fill_template(tmpl_m, list(
  n_treatments       = 4,
  n_replicates       = 5,
  n_total            = 20,
  design_description = paste(
    "Treatments (four rice varieties: A, B, C, D) were laid out in",
    "five randomised complete blocks."
  ),
  model_formula      = "y_{ij} = \\mu + \\tau_i + \\beta_j + \\varepsilon_{ij}",
  software_versions  = paste0("R ", R.version$major, ".", R.version$minor),
  citations          = "R Core Team (2024)."
))

cat(filled_methods)
```

### 6.2  Results Template

```{r template-results}
tmpl_r  <- thesis_template_results(include_posthoc = TRUE)
tmpl_r$placeholders
```

```{r fill-results}
anova_tbl <- export_results_table(aov_result, format = "markdown")
means_tbl <- export_results_table(dmrt_result, format = "markdown")

filled_results <- fill_template(tmpl_r, list(
  anova_significance = "significant",
  response_variable  = "grain yield",
  table_number       = "1",
  anova_table        = anova_tbl,
  min_mean           = round(min(variety_means), 2),
  max_mean           = round(max(variety_means), 2),
  means_letters      = means_tbl,
  figure_number      = "1",
  figure_caption     = "Boxplot of grain yield by variety.",
  interpretation     = paste(
    "Variety C produced the highest yield while Variety D was the lowest.",
    "All pairwise differences were significant at P <= 0.05."
  )
))

cat(filled_results)
```

### 6.3  Discussion Template

```{r template-discussion}
tmpl_d <- thesis_template_discussion(
  topic               = "grain_yield",
  include_limitations = TRUE
)

filled_disc <- fill_template(tmpl_d, list(
  key_findings        = "Significant variety effects on grain yield were observed (P < 0.05).",
  comparison_literature = "These results are consistent with previous reports of variety C superiority (Smith et al., 2020).",
  implications        = "Variety C should be recommended for adoption in similar agro-ecological zones.",
  limitations         = "The trial was conducted in a single season; multi-season validation is needed.",
  conclusion          = "Variety C significantly outyielded all other entries in this study."
))

cat(filled_disc)
```

---

## 7  Compiling a Complete Report

`compile_report()` assembles all components into a single Rmd document and
renders it if **rmarkdown** is available.

```{r compile, eval=FALSE}
res <- compile_report(
  methods_text_obj = mt,
  results_tables   = list(
    "ANOVA Table"  = export_results_table(aov_result,  format = "markdown"),
    "Means Table"  = export_results_table(dmrt_result, format = "markdown")
  ),
  figures          = c("yield_boxplot.png"),
  appendix_obj     = app,
  output_file      = "rice_yield_report",
  template         = "html",
  title            = "Rice Variety Yield Trial — Statistical Analysis",
  author           = "A. Researcher",
  date             = format(Sys.Date(), "%d %B %Y")
)

cat("Report written to:", res$output_file, "\n")
cat("Sections included:", paste(res$sections_included, collapse = ", "), "\n")
```

---

## Summary

The Module 6 workflow in agristat follows these steps:

1. **Design** — create a layout with `design_rcbd()` etc.
2. **Analyse** — run `analyze_design()` and post-hoc tests.
3. **Methods text** — `generate_methods_text()` → paste into thesis.
4. **Tables** — `export_results_table()` in the format your document needs.
5. **Figures** — `export_figure()` for publication-quality images.
6. **Appendix** — `generate_appendix()` for supporting materials.
7. **Templates** — use `thesis_template_*()` + `fill_template()` for a
   consistent academic writing style.
8. **Report** — `compile_report()` to bundle everything automatically.
